TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-07-24 paper/draft.tex
change final # sources

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
with their properties and compositions.
However, insufficient physical assumptions behind atmospheric models
have limited abilities at reproducing spectral features in stars of the 
coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and fully
exploiting the information content and statistical extent of
current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities,
spectral types for 10,311 M dwarfs in the \apogee\ survey using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs
with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
0.9 types.
\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
apply our models and compare our \cannon\ derived $\teff$ against color--temperature
relations and \cannon\ derived $\feh$ against isochrone models in
\gaia\ color-magnitude space.
Finally we use our results to
estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
\end{abstract}

@@

Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of single, main-sequence M stars. Although our catalog reports the temperatures, metallicities, and spectral types for 10,311 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

\color{gcolor}{As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures). To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and ASPCAP vs. photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.

@@

% \begin{figure}
% 	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space. The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr). In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
% 	\label{fig:isochrone_cmd}
% \end{figure}

% \begin{figure*}
% 	\plotone{figures/aspcap_cannon_label_hist.png}
% 	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline. Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
% 	\label{fig:aspcap_cannon_label_hist} 
% \end{figure*}

@@

\begin{figure*}[h]
	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_teff_safe.png} \\
	\includegraphics[width=.45\linewidth]{figures/cmd_feh_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_feh_safe.png} \\
	\includegraphics[width=.45\linewidth]{figures/cmd_spt_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_spt_safe.png}
	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs). Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels. Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected. Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
	\label{fig:safe_selection}
\end{figure*}

@@

We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
from high-resolution infrared spectra from \apogee.
This work was motivated by the problem that M dwarfs stars are difficult to
model physically; the data are better than the models in important senses.
And indeed we find that our data-driven model is both accurate in the
data domain (as a spectral synthesis model) and precise in the latent domain
(as a tool for deriving physical parameters).
This accuracy and precision is consistent with previous work with
\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
extended to a new regime in spectral type ($\teff$) and surface
gravity.
The primary result of this work is that
we have compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures,
metallicities, spectral types, and six-dimensional kinematics.
These data are provided in Table \ref{table:test_results}.

>>>>

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
with their properties and compositions.
However, insufficient physical assumptions behind atmospheric models
have limited abilities at reproducing spectral features in stars of the 
coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and fully
exploiting the information content and statistical extent of
current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities,
spectral types for 5,875 M dwarfs in the \apogee\ survey using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs
with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
0.9 types.
\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
apply our models and compare our \cannon\ derived $\teff$ against color--temperature
relations and \cannon\ derived $\feh$ against isochrone models in
\gaia\ color-magnitude space.
Finally we use our results to
estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
\end{abstract}

@@

Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of single, main-sequence M stars. Although our catalog reports the temperatures, metallicities, and spectral types for 5,875 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

\color{gcolor}{As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures). To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and ASPCAP vs. photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.

@@

% \begin{figure}
% 	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space. The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr). In the ASPCAP plot only the 8,960 out of 5,875 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
% 	\label{fig:isochrone_cmd}
% \end{figure}

% \begin{figure*}
% 	\plotone{figures/aspcap_cannon_label_hist.png}
% 	\caption{Distribution of temperatures and metallicities for 5,875 sources from \thecannon\ compared to the ASPCAP pipeline. Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 5,875 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
% 	\label{fig:aspcap_cannon_label_hist} 
% \end{figure*}

@@

\begin{figure*}[h]
	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_teff_safe.png} \\
	\includegraphics[width=.45\linewidth]{figures/cmd_feh_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_feh_safe.png} \\
	\includegraphics[width=.45\linewidth]{figures/cmd_spt_full.png}
	\includegraphics[width=.49\linewidth]{figures/cmd_spt_safe.png}
	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs). Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels. Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected. Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
	\label{fig:safe_selection}
\end{figure*}

@@

We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
from high-resolution infrared spectra from \apogee.
This work was motivated by the problem that M dwarfs stars are difficult to
model physically; the data are better than the models in important senses.
And indeed we find that our data-driven model is both accurate in the
data domain (as a spectral synthesis model) and precise in the latent domain
(as a tool for deriving physical parameters).
This accuracy and precision is consistent with previous work with
\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
extended to a new regime in spectral type ($\teff$) and surface
gravity.
The primary result of this work is that
we have compiled a catalog of 5,875 M dwarfs with \cannon\ temperatures,
metallicities, spectral types, and six-dimensional kinematics.
These data are provided in Table \ref{table:test_results}.
